---
title: "Analysis of single cell datasets"
output: html_document
---


# Loading libraries

```{r}
library(Seurat)
library(uwot)
library(Matrix)
library(rafalib)
library(dplyr)
remotes::install_github("czarnewski/niceRplots")
library(niceRplots)
library(harmony)
library(biomaRt)
```


# Bone marrow from Tabula Muris (GSE109774, healthy)

## Downloading datasets from GEO

```{r}
# CREATE FOLDER STRUCTURE
PATH <- "../data/tabula_muris_GSE109774/"
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc="
dir.create(PATH, recursive = T)


# DOWNLOAD THE TWO BONE MARROW SAMPLES
download.file(
  url = paste0(url,"GSM3040900&format=file&file=GSM3040900%5FMarrow%2D10X%5FP7%5F2%2Etar%2Egz"),
  destfile = paste0("../data/tabula_muris_GSE109774/GSM3040900.tar.gz"))
download.file(
  url = paste0(url,"GSM3040901&format=file&file=GSM3040901%5FMarrow%2D10X%5FP7%5F3%2Etar%2Egz"),
  destfile = paste0("../data/tabula_muris_GSE109774/GSM3040901.tar.gz"))


# DECOMPRESS FILES
untar(tarfile = "../data/tabula_muris_GSE109774/GSM3040900.tar.gz",
      exdir = "../data/tabula_muris_GSE109774/")
untar(tarfile = "../data/tabula_muris_GSE109774/GSM3040901.tar.gz",
      exdir = "../data/tabula_muris_GSE109774/")


# CLEAN THE DIRECTORY
file.remove(c("../data/tabula_muris_GSE109774/GSM3040900.tar.gz",
              "../data/tabula_muris_GSE109774/GSM3040901.tar.gz") )
```


## Basic analysis using Seurat

```{r}
# CREATE SEURAT OBJECT FOR EACH DATASET
dir_list <- list.dirs( PATH , full.names = F)[-1]
rawdata <- lapply( dir_list , function(x) {
  CreateSeuratObject(counts = Read10X(paste0(PATH,x)) , project = x) } )
names(rawdata) <- dir_list


# MERGE OBJECTS INTO A SINGLE BONE_MARROW
rawdata[["bone_marrow"]] <- merge(
  x = rawdata[["Marrow-10X_P7_2"]],
  y = rawdata["Marrow-10X_P7_3"])
rawdata <- rawdata[-c(1:2)]


# PERFORM STANDARD ANALYSIS USING SEURAT
# Key function parameters used are shown (defaults were used otherwise)
rawdata[["bone_marrow"]] <- NormalizeData( rawdata[["bone_marrow"]] )
rawdata[["bone_marrow"]] <- FindVariableFeatures( rawdata[["bone_marrow"]],nfeatures = 4000 )
rawdata[["bone_marrow"]] <- ScaleData( rawdata[["bone_marrow"]] ,vars.to.regress = "nFeature_RNA")
rawdata[["bone_marrow"]] <- RunPCA( rawdata[["bone_marrow"]] ,npcs = 100 )
rawdata[["bone_marrow"]] <- RunUMAP(
  object = rawdata[["bone_marrow"]] ,dims = 1:50, reduction = "pca",
  spread = .5,min.dist = 0.3, repulsion.strength = 0.4 )
rawdata[["bone_marrow"]] <- FindNeighbors( rawdata[["bone_marrow"]], dims = 1:30, reduction = "pca" )
rawdata[["bone_marrow"]] <- FindClusters( rawdata[["bone_marrow"]], resolution = 1 )
plot_meta(rawdata[["bone_marrow"]],red="umap",feat = "seurat_clusters",cex = 0.5)
plot_meta(rawdata[["bone_marrow"]],red="umap",feat = "orig.ident",cex = 0.5)
rawdata[["bone_marrow"]]$cell_name <- paste0("BM_",rawdata[["bone_marrow"]]$seurat_clusters)


# SAVE THE SEURAT OBJECT
saveRDS(rawdata[["bone_marrow"]],"../results/bone_marrow.rds")
```


# Create a joint UMAP for all datasets

## Merge all datasets into 1 SeuratObject

```{r}
# LOAD DATASETS
rawdata <- list()
rawdata[["bone_marrow"]] <- readRDS("../results/bone_marrow.rds")
rawdata[["bcell"]] <- readRDS("../results/Bcell.combined.RDS")
rawdata[["epi"]] <- readRDS("../results/epi.int.RDS")
rawdata[["stromal"]] <- readRDS("../results/stromal.int.RDS")


# FIX THE CELL NAMES FOR COHERENCE
rawdata[["bcell"]]$cell_name <- rawdata[["bcell"]]$celltype.stim
rawdata[["bcell"]]$cell_name <- paste0( "B_", rawdata[["bcell"]]$stim,"_",rawdata[["bcell"]]$celltype )
rawdata[["stromal"]]$cell_name <- paste0( "S_",gsub("depletion","BCD",gsub("control","CTRL",(rawdata[["stromal"]]$old.ident))), "_", rawdata[["stromal"]]$integrated_snn_res.0.2  )
rawdata[["epi"]]$cell_name <- paste0( "E_",gsub("depletion","BCD",gsub("control","CTRL",(rawdata[["epi"]]$old.ident))),"_",rawdata[["epi"]]$integrated_snn_res.0.3 )


# MERGE OBJECTS
data <- merge(x = rawdata[[1]],y=rawdata[-1])
```


## Define matrix rotation function

In linear algebra, a rotation matrix is a transformation matrix that is used to perform a rotation in Euclidean space. For example, using the convention below, the matrix:

$$\begin{bmatrix}cos\theta & -sin\theta\\ sin\theta & cos\theta \end{bmatrix}$$

```{r}
rotate <- function(mat,angle){
  angle <- ((angle)/180)*pi
  rot_mat <- matrix(c(cos(angle),sin(angle),-sin(angle),cos(angle)),nrow = 2,byrow = T)
  return(mat %*% rot_mat)
}
```


## Compute the joint UMAP coordinates

```{r}
# Re-scale the individual umap embeddings [0-1]
norm1 <- apply(rawdata[["bone_marrow"]]@reductions$umap@cell.embeddings,2,function(x) (x-min(x))/(max(x)-min(x)) )
norm2 <- apply(rawdata[["bcell"]]@reductions$umap@cell.embeddings,2,function(x) (x-min(x))/(max(x)-min(x)) )
norm4 <- apply(rawdata[["epi"]]@reductions$UMAP_on_integrated@cell.embeddings,2,function(x) (x-min(x))/(max(x)-min(x)) )
norm5 <- apply(rawdata[["stromal"]]@reductions$UMAP_on_integrated@cell.embeddings,2,function(x) (x-min(x))/(max(x)-min(x)) )

# Rotate the unit vector in a circle
unit_vector <- c(0,1)
res1 <- rotate(unit_vector,45+90*0)
res2 <- rotate(unit_vector,45+90*1)
res4 <- rotate(unit_vector,45+90*2)
res5 <- rotate(unit_vector,45+90*3)

# Compute the new coordinated for each dataset
norm1t <- norm1 + matrix(rep(res1,nrow(norm1)),ncol = 2,byrow = T)
norm2t <- norm2 + matrix(rep(res2,nrow(norm2)),ncol = 2,byrow = T)
norm4t <- norm4 + matrix(rep(res4,nrow(norm4)),ncol = 2,byrow = T)
norm5t <- norm5 + matrix(rep(res5,nrow(norm5)),ncol = 2,byrow = T)

# Create new UMAP layout
joint_umap <- rbind(norm1t,norm2t,norm4t,norm5t)
```

## Inspect joint UMAP using convex-hulls

This is to make sure there are no cells in displayed in the wrong UMAP area.

```{r}
plot(joint_umap,cex=.3,type="n")
ch <- chull(norm3)
polygon( norm3[c(ch,ch[1]),1] , norm3[c(ch,ch[1]),2] , col = "grey90", border = "grey80")
ch <- chull(norm1t)
polygon( norm1t[c(ch,ch[1]),1] , norm1t[c(ch,ch[1]),2] , col = "grey90", border = "grey80")
ch <- chull(norm2t)
polygon( norm2t[c(ch,ch[1]),1] , norm2t[c(ch,ch[1]),2] , col = "grey90", border = "grey80")
ch <- chull(norm4t)
polygon( norm4t[c(ch,ch[1]),1] , norm4t[c(ch,ch[1]),2] , col = "grey90", border = "grey80")
ch <- chull(norm5t)
polygon( norm5t[c(ch,ch[1]),1] , norm5t[c(ch,ch[1]),2] , col = "grey90", border = "grey80")
points(joint_umap,cex=.1)
```

## Add joint UMAP to the SeuratObject and save

```{r}
# ADD JOINT UMAP AND PLOT PER CELL CLUSTER
data@reductions[['joint_umap']] <- CreateDimReducObject(embeddings = joint_umap)
plot_meta(data,"joint_umap","cell_name",cex=.5)


# SAVE OBJECTS AND NORMALIZED MATRICES
saveRDS(data,"reference_scRNAseq_colon_BM.rds")
save_matrix_to_HDF5( data@assays$RNA@data , "colonic_reference_scRNAseq_normcounts.h5" )
```
